\chapter{Conclusions}
\label{chap:conclusions}
% gancho para motivação e objetivos; introduir o assunto com um pouco mais de profundidade

% 1) enfatizar contribs
% --> falar dos resultados especificose limitaçẽis do experimento; aqui tem que generalizar se possível
% 2) falar de limitações e dificuldades
% 3) trabalhos futuros
% 4-8 de julho

% amarração do Data literacy; o experiemento teve a ver com data litearcy
% enfatizar nas limitações que o grupo do DL deveria ter sido o teste stodap

The open data promises are still far from being materialised.
However, it cannot be denied that significant advances have been made in recent years.
In this chapter, the conclusions of this thesis will be derived, based on the initial hypothesis and in the contributions that were made during this work.
Limitations and difficulties found on the thesis development will also be discussed.
Finally, we point out various directions for researchers willing to continue this work.


\section{Contributions}

As described in \autoref{sec:objectives}, this thesis has a main objective, and also other specific ones.
In this section, we first detail the main contribution, analysing the validation of our initial hypothesis and the possible generalizations of our results.
The second part describes other contributions related to the specific objectives.

\subsection{Main Contribution}
\label{sec:main_contribution}

In the Introduction of this work, a hypothesis was posed, which, for the ease of reading, is reproduced here:

\noindent\textbf{H\ref{hyp:tagging}:} \emph{Enhancing the organization of open data repositories leads to better find ability of open data.}


The first chapters of this thesis emphasized the importance of open data in our society, and we showed by literature review (\autoref{sec:problems}), participatory research (\autoref{dl_results}) and objective metrics (\autoref{sec:analysis}) that organization of open datasets is a relevant problem.
The STODaP approach proposed in \autoref{chap:tagging} targeted precisely the open data organization problem, both from a local perspective (inside a single ODP) and from a global perspective (inter-ODPs).
The implementation of STODaP server was evaluated in the previous chapter, and results shown that the system enabled people to find datasets preciser and quicker than using other approaches.
This can be considered true specially for non-experts in data processing and open data, and regarding topics for that open datasets are not abundant.

In order to validate the hypothesis, general theoretical and practical analysis were developed, as described above.
However, experiments are intrinsically specific.
Thus, it is necessary to discuss if the results described in this thesis can be generalised, or if they are true only for a specific context.

As discussed in the conclusions of the last chapter, evaluation of STODaP server has two specific design choices that may affect the generality of the results.
The first aspect was the choice of subjects.
Participants of the experiment can be divided in three groups: (i) first year Computer Sciences students, in Brazil; (ii) Semantic Web researchers of various nationalities; (iii) open data practitioners/activists in Brazil.
Even though there are significant variations on the level of data/open data experience among these subjects, all of them share a daily use of internet and a good English proficiency, and are members of the Computer Science community, or very familiar with it, in the case of some activists.

The second design choice was regarding the questions that subjects had to answer: (i) find 2015 open budget data from 5 countries; (ii) find open data about water quality in seven rivers outside Europe; and (iii) find open procurement datasets in 3 different languages.
The choice of the topics was based on the 13 criteria used by the Open Data Index\footnote{The Open Data Index compiles an open data ranking, whose criteria are: National Statistics, Government Budget, Legislation, Procurement tenders, Election Results, National Map, Weather forecast, Pollutant Emissions, Company Register, Location datasets, Water Quality, Land Ownership and Government Spending. For each of these topics, countries are evaluated and receive a score according to openness level.}, in order to guarantee the relevance of the questions on the open data community.

In this context, regarding the choice of subjects, it is not possible to guarantee the extension of our results to subjects completely unaware of the Computer Science field, with a low English proficiency, or with a low computer/internet ability.
Regarding the topics, since all Open Data Index topics have a significant amount of open datasets, it is secure to generalize our findings to other relevant open data topics.
As seen in last chapter, a higher availability of datasets favours generic web search engines.
No conclusions can be drawn about topics less popular topics, but we can speculate that STODaP would have a good performance as long as datasets exist and are adequately tagged.

Thus, we can affirm that:

\noindent \emph{For users with at least an intermediate level of English and some approximation to the Computer Science field, STODaP open data search engine delivers open datasets with a higher precision in less time than other search methods when searching for relevant open data topics.}

\subsection{Other Contributions}
\label{sec:other_contributions}

\autoref{chap:dataliteracy} presents some contributions on the Data Literacy field.
The importance of data literacy efforts was emphasized in \autoref{dl_paulofreire}, where an analogy between traditional and data literacies was derived.
We concluded showing the importance of data literacy on giving voice to marginalized people, in the same way as it was crucial to alphabetise people some decades ago.
A definition of the concept of \emph{Critical Data Literacy} was also presented, emphasizing the need of a real understanding of what is behind data.
In \autoref{dl_method} our proposal for working on data literacy with social movements activists was presented.
A methodology for data literacy course was detailed, mixing theory, discussion and practice, in an effort to bring data literacy knowledge closer to each educands reality.

Thus, on the Data Literacy field, we bring:
\begin{itemize}
	\item A theoretical contribution regarding the contributions of Popular Education theory to Data Literacy, and the definition of the concept of Critical Data Literacy;
	\item A methodological contribution, regarding the description of a Data Literacy course and associated research methodology, with an emphasis on a critical understanding and use of data by social movement activists; and
	\item A practical contribution, regarding the systematisation of impediments, benefits and improvements of open data according to social movement activists.
\end{itemize}

\autoref{chap:relworks} brings also an as a contribution a framework for analysing the use of tags inside ODPs and between several ODPs.
Metrics developed are not innovative themselves, but their application on the ODP context bring lights to the problematic use of tags.

In \autoref{chap:tagging}, beyond the already presented contribution, there are still local strategies for cleaning up and semantically lifting tags.
These contributions -- the Tag Manager Plugin and the Semantic Tags Plugin -- are part of the STODaP approach, but were not evaluated because their use demand modifying ODPs software, which is managed by government servants.

% Our hypothesis H\ref{hyp:dataliteracy} was partially validated in \autoref{dl_results}, where the results of the application of this course were presented.
% Specially, in \autoref{fig:dl_results}, we depicted the critical apprehension of data literacy course participants regarding the problems of open data.
% Tables~\ref{tab:dl_results1}, \ref{tab:dl_results2}, and \ref{tab:dl_results3} complement the diagram and present in a more detailed fashion the outputs of the data literacy course.
% The critical use of data by subjects, however, could not be evaluated.
% As discussed in \autoref{sec:obs_based_analysis}, the fourth stage of the course, were the real use of data was planned, was reached only twice, and thus it would be precipitated to take any conclusions about it.

% Regarding the results achieved by the data literacy experiment, specificity lies in the main characteristic of the participants, i.e., social movements activists.
% Although the open data impediments gathered on the course do not differ substantially from others reported on the literature, we can see that motivations and desired improvements are very specific of social movements activities.

% Thus, we can affirm that:

% \noindent \emph{The developed data literacy course methodology was able to raise critical a critical view about open data on social movement activists. However, the use of open data by theses subjects could not be tested.}



\section{Limitations and Difficulties}

After highlighting the contributions of this thesis, it is also necessary expose some limitations and difficulties, and delimit the extension of our results.

Regarding the work reported in this thesis as a whole, the main conceptual problem lies in attempting to derive a transdisciplinary approach without having the appropriate means for it.
Open Data is a topic that comprehends multiple perspectives.
Political sciences, regarding open data policies, Law Studies, regarding access to information regulation, Social Sciences, regarding the effects it can have on the society, or Education, regarding Data Literacy are certainly a few of them.
Looking at the problem from the Computer Science point of view is necessary, but the other perspectives have to be somehow included.
According to the Charter of Transdisciplinarity, 

\begin{citacao}
The transdisciplinary vision is resolutely open insofar as it goes beyond the field of the exact sciences and demands their dialogue and their reconciliation with the humanities and the social sciences as well as with art, literature, poetry and spiritual experience. \cite[Article 5]{Freitas1994}
\end{citacao}

The initial aim of this thesis was to achieve a balance between social and technical aspects, in the sense of connecting people's real demand to the developed approach, recognizing the complexity and transdisciplinarity of the open data problem.
Although it was still possible to include social and educational aspects, mainly through the Data Literacy discussion, their role ended up being more motivational than really being part of the scientific development.
A reason for this lies on the missing transdisciplinary view on the Computer Sciences field, which hampers the validation of a hypothesis off the objective and quantitative framework, derived from the Cartesian scientific method.

More specifically, by choosing the problem of open data organization, we left several other ones detected in \autoref{chap:dataliteracy} open.
It is of course impossible to deal with every problem on the open data field in a single thesis, but some of them are of crucial importance, as for example the Data Quality problem.
In our evaluation procedure, subjects went only until the entry point of datasets.
It is possible that datasets that were marked as valid in the evaluation do not correspond to their descriptions, are outdated, or have broken links to their resources.
This is a clear limitation of this work: open datasets search is based on metadata, but data quality cannot be guaranteed.

Another limitation of the approach is related to the subjects that participated on the evaluation.
Ideally, we would expect that the same subjects that attended to the Data Literacy course could evaluate the system.
However, this was not possible, mainly because: (i) Most of the participants on the Data Literacy course do not speak English; (ii) the time span between the courses and the evaluation was of almost 2 years, which makes contacts more difficult.
Thus, it was not possible to evaluate if the developed system attended to at least some of the requirements pointed out by participants of the courses.

Moreover, the connection between tags and semantic tags is still limited, as mentioned previously.
The automatic approach used in this thesis resulted in many tags not connected to right semantic tag, and also many wrong connections. 
Although several methods to tackle this issue are available in the literature, the large size of our database requires efficient approaches that are still to the tested at STODaP.

Instability of Open Data Portals was another limiting factor.
As reported in \autoref{chap:relworks}, from 140 listed portals, only 87 could be accessed.
Thus, our database could be almost twice as complete as it is now.
However, due to several reasons stated before, ODP could not be accessed.

\section{Future Works}

The broadness of this work let many open paths for further research.

On the Data Literacy field, \autoref{fig:dl_results} can give relevant clues for further research.
The training branch presents a great variety of topics, ranging from Basic Informatics, Mathematics and Statistics, until data journalism or open data publishing.
The possibility of integrating such a wide range of topics in Data Literacy courses should be further investigated.

Regarding reconciliation of tags in ODPs, a sort of crowd validation would be very useful in order to confirm automatic processing and add new information.
The model proposed by \citeonline{Limpens2013} could be very useful, because of its ability to deal with divergences between users.
In this sense, it is necessary to call the attention of the open data community in order to further advance on collaborative strategies for enriching the semantic tags server.
Another way of improving the connection between tags and groups with their semantic equivalents could be via artificial intelligence methods. 

Advances on local tools for tagging assistance are crucial, since a good tagging procedure would certainly ease the work of tag reconciliation. 
Future research and development should include a tag suggestion approach for ODPs which takes into account the related tags at the tag server, using collective knowledge as in~\citeonline{Sigurbjornsson2008}.
Using the possibly structured data of the ODPs in order to improve tagging suggestions is also a research direction that should be followed.

Another interesting research direction is the detecting the emergence of schemas from the tags, as described in~\citeonline{Robu2009}, where tag correlation is used to create tagging vocabularies and visualizations of terms relationships.
A schema emerged from ODP content and metadata could facilitate users understanding of the portal organization.

The STODaP system also need a more professional approach on its development in order to be really useful. A priority list of enhancements could start by: 

\begin{itemize}
	\item \emph{Internationalization.} At the moment, semantic tags and groups labels are only in English. This means that users searching for keywords in English are able to find datasets in other languages via semantic tags, but the contrary is not true. The Gemet Thesaurus offers several translations of the terms, which could be used for this task.
	\item User Interface Design. Several subjects commented on the evaluation that interface design should be improved. This is a priority because interface problems can hide system functionalities, and thus result in a worse performance than the system could offer.
	\item Include other open data portals. The list of ODPs indexed by the STODaP server came from the CKAN Census. The first problem is that this Census is not frequently updated, since many CKAN ODPs listed there do not exist more, and other not present in the Census were found. Secondly, a great quantity of ODPs are developed using Socrata, another open data management system. Although being proprietary, Socrata also offers APIs for external users to consume data. Developing the interface between CKAN and Socrata could significantly improve the STODaP base.
\end{itemize}

\vspace{1cm}
\begin{center}* * *
\end{center}
\vspace{1cm}

For STOdAP to realize its full potential, ODP administrators and users should be involved and (meta)data literacy needs to be improved.
Far from being a technical problem, open data will not advance if politicians and public servants in charge of transparency policies do not give special attention to data literacy.
Including critical data literacy on scholar curriculum could consistently enhance the level of data skills on the society.

Although open data in currently a trendy word, its ``openness'' is still limited by many factors, including political and technological ones.
With this work, we hope to have contributed to the open data field, maybe not so much with the resulting tool, but hopefully with a socio-technical view where the user side can have at least the same importance as the publisher side.